---
title: "Cellosaurus"
author: "Michael Steinbaugh"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{Cellosaurus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE}
## nolint start
suppressPackageStartupMessages({
    library(DT)
    library(Cellosaurus)
})
## nolint end
```

# Introduction

The package is designed to import [Cellosaurus][] annotations and serve as a
cell line identifier mapping toolkit. The main `Cellosaurus()` generator
function returns a `Cellosaurus` object, which extends the S4 `DFrame` class
from [Bioconductor][]. Additional functions are defined to work with the
`Cellosaurus` object, notably `mapCells()`, which provides mapping support to
Cellosaurus identifiers directly from cell line names, [DepMap][] identifiers,
and [ATCC][] identifiers.

# Import Cellosaurus table

The primary generator functions imports annotations from the `cellosaurus.txt`
file on the [FTP server][].

```{r import, message=FALSE}
object <- Cellosaurus()
print(object)
```

The object is structured as cells in rows, metadata in columns.

```{r}
print(colnames(object))
```

Data is encoded using run-length encoding (`Rle` from the S4Vectors package) to
lower memory overhead. 

This approach provides simple spreadsheet-like access to Cellosaurus
annotations, which are more intuitive to users than nested JSON-style lists.

```{r}
i <- seq(from = 1L, to = 10L)
j <- which(vapply(
    X = as.data.frame(cello),
    FUN = is.atomic,
    FUN.VALUE = logical(1L)
))
datatable(
    data = as.data.frame(object)[i, j],
    options = list(scrollX = TRUE)
)
```

# Mapping cell lines

It remains a common problem in cancer research that cell line inventories are
only maintained by cell line name, without any systematic standardization
against a reference database, such as Cellosaurus, ATCC, or DepMap. The
advantage of standardizing upon Cellosaurus identifiers as the primary
[research resource identifier][rrid] is that the Cellosaurus database is
a superset of cells that include all cells in ATCC and DepMap. The Cellosaurus
database also provides nicely curated metadata on problematic cell lines,
notably contamination, and common misspellings of cell line names that persist
across a number of commercial vendor databases.

The `mapCells()` function in the package aims to simplify mapping of cell line
names and identifiers from other databases, notably DepMap and ATCC, to
Cellosaurus identifiers. The function is designed to be as simple as possible
and support mixed input in a single call.

```{r mapping-cell-lines}
cells <- c("CCL-228", "Duadi", "MV411", "U87MG", "RAW 264.7")
i <- mapCells(object, cells = cells)
print(i)
datatable(
    data = as.data.frame(object)[i, j],
    options = list(scrollX = TRUE)
)
```

# Excluding problematic cell lines

The Cellosaurus database nicely keeps track of known issues with cell lines,
which can be broken down roughly into two classes: "problematic" and
"contaminated". It can be useful for downstream analysis to exclude these
cell lines -- in particular, we recommend generally excluding any contaminated
cell lines but not necessarily all problematic cell lines.

```{r exclude-problematic-cells}
subset <- excludeProblematicCells(object)
print(nrow(subset))
print(any(subset[["isProblematic"]]))
print(any(subset[["isContaminated"]]))
```

Note that "contaminated" cell lines are a subset of "problematic" cell lines in
the database. These in general should be avoided in downstream workflows.

```{r exclude-contaminated-cells}
subset <- excludeContaminatedCells(object)
print(nrow(subset))
print(any(subset[["isContaminated"]]))
print(any(subset[["isProblematic"]]))
```

For reference, search the website for "problematic" cell lines with
`"Problematic cell line"` and "contaminated" cell lines with
`"Problematic cell line: Contaminated"`.

# R session information

```{r session-info}
utils::sessionInfo()
```

[atcc]: https://www.atcc.org/ "American Type Culture Collection"
[bioconductor]: https://bioconductor.org/
[cellosaurus]: https://www.cellosaurus.org/
[depmap]: https://depmap.org/
[ftp server]: https://ftp.expasy.org/databases/cellosaurus/
[rrid]: https://www.rrids.org/ "Research Resource Identifier"
